A multi-scale story of the diffusion of a new technology: the web

### 2025-07-29 paper/submission3/title.tex
submission 3

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}

@@

\affiliation[1]{organization={University of
Bristol},postcode={UK},postcodesep={}}

@@

\section{Introduction}\label{sec-introduction}

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international, ding2010modeling}. As a consequence,
relevant studies did not investigate diffusion mechanisms at granular
geographical scales to provide local policy insights, for instance,
regarding infrastructural development and local technology policies. A
potential explanation of the lack of geographical studies exploring the
diffusion of new and, more specifically for this paper, digital
technologies across both space and time can be attributed to the
scarcity of relevant and granular enough data, a problem also
highlighted by \citet{iso2005} and \citet{kemeny2011international}. As
\citet{zook2022mapping} pointed, digital activities are hardly ever
captured in official data.

This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of a
digital technology that is the Web in the UK at a high level of spatial
granularity from the very early days until its maturity (1996-2012). It
specifically focuses on the diffusion of commercial websites. It does so
by employing a novel source of big data and machine learning algorithms
to model the \emph{active} engagement with Web -- that is creating and
maintaining instead of just browsing a website -- during that period and
tests how different spatial and temporal mechanisms shaped its diffusion
\citep{tranosuk}. This paper also exemplifies how the combination of
data sources and state-of-the-art research methods which escape the
traditional social science domain can offer new lenses to geographical
research regarding the understanding of technological diffusion.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers, network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there is a scarcity of studies analysing the diffusion
of new technologies at a detailed geographical level because of the lack
of detailed enough data and adequate research methods.

This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
\citep{solow1957technical, aghion1990model, kemeny2011international}.
Linked to the subject of this paper, the Web and, more specifically,
commercial websites can be considered as the underpinning technology for
the transformative process identified by \citet{capello2024nexus} as the
\emph{digital service economy}. Importantly, the focus here is not the
Web in general, but instead commercial websites. After all, diffusion
together with invention and innovation are considered the pillars of
technological change \citep{das2022diffusion}. Findings from previous
studies illustrated that early engagement with digital technologies such
as the Web led to longstanding positive productivity gains for these
regions that hosted early adopters \citep{tranosuk}. Hence, obtaining a
robust understanding of the spatial diffusion mechanisms is crucial in
mapping future development trajectories.

Another line of motivation has to do with the very nature of the Web as
a digital technology and how it evolved over time. The digital
geographies field has debunked premature claims about the Internet's
placeless-ness by mapping various of its facets and technological layers
\citep{tranos2013death}, but the lack of relevant long time-series data
prevented researchers from observing digital phenomena overtime. Most of
these studies focused on its hard infrastructure
\citep[e.g.][]{malecki2002economic, moss2000internet}, on Internet
subscribers \citep[e.g.][]{blank2018local}, or on social media users
\citep[e.g.][]{crampton2013beyond}. Exceptions include early work from
\citet{zook2000web}, \citet{zook2001old} and \citet{moss1997tracking},
which utilised domain name registration data and use it to map and
illustrate the spatial footprint of the web content. Pivotal as these
contributions have been in understanding the economic geographies of the
Internet, they were based on cross-sectional, often spatially detailed,
mapping and on short temporal comparisons (e.g.~1993-1998) and did not
observe the evolution and diffusion of the Web until its maturity. As we
know from the relevant literature, different technologies diffuse at
different rates and there is no theory that can explain these
differences \citep{leibowicz2016representing}. Hence, mapping and
modelling the diffusion of the Web can offer useful insights.

The lack of previous granular studies about the diffusion of new
technologies deprives us from a reliable prior regarding whether the
diffusion mechanisms and drivers are sensitive or not to the spatial
scale that the diffusion of a new technology is observed. As per other
key concepts in economic geography
\citep{briant2010dots, meliciani2015determinants, andini2019social, rosenthal2001determinants},
diffusion mechanisms can indeed suffer from Modifiable Areal Unit
Problem (MAUP): although the adoption of a new technology is at its core
an individual process, it is hardly ever observed at this, or at a close
enough, scale.

This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
Internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.\footnote{\url{https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft}}
Also, only 9 per cent of UK's households had access to the Internet in
1998 \citep{ons2018}, the web included mostly static webpages, there
were no social media and web browsing involved exclusively a desktop PCs
as there were no smartphones \citep{tranosuk}. Hence, it is fair to say
that the study period captures the very early stages of the diffusion of
a new technology that is the Web until its maturity. The former is a key
point in the lifecycle of a new technology. Firstly, during this period
new technologies are expensive, crude and imperfect
\citep{rosenberg1994exploring, wilson201281}. A simple comparison
between Web 1.0 and Web 3.0 applications clearly illustrates this
\citep{tranos2020social}: while a static website facilitates one-way
information dissemination, a platform like \texttt{github} enables
cooperation between users and the creation of new information, meaning,
and even knowledge \citep{faraj2016special, barassi2012does}. During
this period the performance of a new technology is the main attraction
and not the cost to access and use it \citep{wilson2011lessons}. There
is a broader theoretical discussion in the literature about the early
adoption motives. As summarised by \citet{perkins2005international}, on
one hand, epidemic models highlight the role of interpersonal contacts
as a way for new technologies to diffuse. On the other hand, economic
models underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories \citep{neffke2011regions, simmie2014new}. This
argument is even more relevant when the focus is on digital technologies
because of the commonly found lag between investment and economic
returns as reflected in the Solow paradox
\citep{acemoglu2014return, brynjolfsson2018artificial}.

The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the Internet such as Internet adoption and speeds
\citep[e.g.][]{blank2018local, destefano2022fuzzy}, this paper observes
yearly counts of geolocated, commercial websites, which reflect the
creation and maintenance of such websites. It needs to be highlighted
here that although there are various types of underpinning web
technologies -- namely markup, stylesheet, client- and server-side
coding languages as well as different web browsers and other related
programming language -- the focus here is on the diffusion of the Web as
an overarching technology instead of more specific web technologies. For
a such a detailed and cross-country technological diffusion study, see
\citet{PAPAGIANNIDIS2015308}.

To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. The exploratory
spatial data analysis offers some first indications of the importance of
the different spatial diffusion mechanisms in shaping the diffusion of
the Web in the UK. Importantly, the analysis adopts two distinct
geographical scales: the LAD and the most detailed scale of UK census
areas, the Output Areas (OA). Then, a novel methodological framework
based on Machine Learning (ML) algorithms is developed to test the role
of such mechanisms at these two scales. The results reveal, for the
first time, the importance of different spatial diffusion mechanisms in
the diffusion of a digital technology at very granular geographical
scales. The analysis also reveals that the diffusion of the Web is
characterised by spatial heterogeneity, well-established spatial
patterns and both stability and volatility.

The structure of the paper goes as follows. Section~\ref{sec-litreview}
reviews the mechanisms that shape the diffusion of a new technology and
sets up the two hypotheses that are empirically tested in
Section~\ref{sec-results}. Section~\ref{sec-datamethods} describes the
data and methods used in this paper. The discussion of the results and
the conclusions of the paper are offered in
Section~\ref{sec-conclusions}.

\section{The mechanisms that shape the diffusion of a new
technology}\label{sec-litreview}

The diffusion of new technologies and innovations is driven by three
different mechanisms \citep{grubler1990rise, morrill2020spatial}: (i)
distance and proximity, (ii) urban hierarchies in the form of a
centrifugal forces as new technologies diffuse from core to periphery,
and (iii) over time the cumulative level of adoption follows an S-shaped
pattern. Although some studies focused on other, case-specific drivers
-- for example \citet{lengyel2020role} on urban scaling and
\citet{bednarz2020pulled} on regional demand -- the literature is in
agreement that these are the three fundamental diffusion mechanisms.
\citet{hagerstrand1968innovation} was the first to identify diffusion as
a geographical process. The starting point was the idea that diffusion
is based on passing information through social networks, which
themselves tend to be geographically bounded. Hence, he identified a
`neighbourhood' mechanism of how information, and consequently,
innovations and new technologies diffuse. This perspective draws
similarities with epidemics: new technologies and innovations, just like
pathogens, spread because of contagion and, consequently, proximity and
exposure \citep{hivner2003facilitating}.

\citet{hagerstrand1968innovation} also emphasised the role of the
top-down hierarchy of urban systems and highlighted that new
technologies and innovations are firstly adopted in larger cities and
then diffuse to second tier ones. This is a sequential instead of a
simultaneous process, which resembles the `lead-lag' spatial
acceleration effect in market research
\citep{bento2018time, PERES201091}. Schmidt's Law empirically
illustrates a similar pattern \citep{grubler1990rise}. Core and highly
agglomerated regions is where new technologies are invented and
commercially deployed. This is where the first adopters tend to be
based. Then, technologies spread to the rim and eventually to the
periphery. Although adoption pace might be higher when new technologies
finally arrive to the periphery, the saturation levels may never reach
the ones in the core because of the lack of infrastructure or other
necessary institutions \citep{leibowicz2016representing}.

Diffusion is naturally a dynamic process and Hägerstrand recognised the
role time plays in the diffusion of innovations and new technologies: an
early-pioneering period, a middle fast accelerating period and a final
saturation one \citep{morrill2020spatial}. This was further explored by
\citet{rogers2010diffusion}, who identified five types of adopters of
new technologies and innovations: innovators, early adopters, early
majority, late majority and laggards. Hence, plotting the cumulative
adoption of a new technology over time often leads to an S-shaped line
plot \citep{grubler1990rise}.

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge
gap regarding the explanatory power of these mechanisms at granular,
local scales \citep{ding2010modeling}. As discussed in
Section~\ref{sec-introduction}, such spatially explicit knowledge can
directly inform local technology and infrastructure policies. Indeed,
while the spatial dimension is present in the following studies, the
geographical level of analysis is often too coarse to ascertain the role
of the above mechanisms at the local scale. \citet{beardsell1999spatial}
studied the evolution of the computer industry in 317 US metropolitan
areas during the 1977-1992 period using employment data. Their analysis
indicated that the relative size distribution holds for urban computer
employment and also urban heterogeneity is essential in explaining this
distribution. In a recent study, \citet{bednarz2020pulled} focused on
wind turbines and modelled their spatial diffusion across 402 German
regions during 1970-2015. Their key finding is that local demand than
local supply was the main driving factor. \citet{haller2011determinants}
employed firm panel data for Ireland during 2003-2005 and illustrated
that when firms are located in a region or industry with high share of
firms having a website, they are more likely to have a website too. At a
global scale \citet{perkins2011internet} explored whether the adoption
of previous communication technologies that is mail, telegrams and
telephones were shaped by similar socioeconomic factors as the Internet.
Their results indicated common patterns regarding the drivers behind the
adoption of different communication technologies.
\citet{feldman2015rknowledge} focused on inventions and used patent data
to map how the the rDNA technology diffused among inventors across 366
US metropolitan areas. \citet{ding2010modeling} modelled the spatial
diffusion of mobile telecommunications across the highest level of
administrative division in China (29 regions in total). Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
\citet{bakher2013diffusion} compared different growth curves as a means
to model the diffusion of online shopping in Australia during the
1998-2009 period, but they did not consider any spatial mechanisms. At a
global scale \citet{PAPAGIANNIDIS2015308} modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
\citet{lengyel2020role}, who analysed the adoption and churn of a
Hungarian online social network. Their results were in agreement with
the early theoretical and empirical contributions reviewed here:
assortativity, urban scaling and distance are the key drivers of spatial
diffusion. Based on the above, the main empirical hypothesis that is
tested in this paper states:

\begin{quote}
H\textsubscript{1}: The three diffusion mechanisms namely distance,
urban hierarchy and the S-shaped pattern of the cumulative level of
adoption shaped the diffusion of the commercial Web in the UK even at
granular geographical scales.
\end{quote}

Understanding the diffusion of a new technology in multiple, detailed
spatial scales is of interest to economic geography as potential local
economic benefits may occur because of concentration of early adopters
or through technological leapfrogging. On the one hand, early adopters
are often rewarded because of their attitudes towards new technologies
and innovations. \citet{rogers2010diffusion} identifies them as
`knowledgeable risk takers' and \citet{griliches1957} as `profit
maximisers' \citep{ding2010modeling}. On the other hand, latecomers can
adopt and benefit from new technologies that have been developed
elsewhere without incurring the hefty initial R\&D costs
\citep{teece2008firm}. At a global scale
\citet{perkins2005international} explored whether the diffusion rate of
new technologies is driven by a latecomer advantage and the engagement
with the global economy via foreign direct investments and trade. Their
results illustrate that indeed latecomers and developing countries
experience diffusion of new technologies more rapidly than early
adopters and developed countries. At a regional scale, previous research
highlighted the long term and sustained productivity benefit of the
early adopters of web technologies in the UK \citep{tranosuk}. Although
the early adoption and leapfrogging literature has not paid much
attention on cities and regions \citep{yu2018sustainability}, the
potential local economic benefits justify the need to investigate the
dynamics of the web diffusion in different scales in the UK. Hence, the
second empirical hypothesis states:

\begin{quote}
H\textsubscript{2}: The diffusion of the commercial Web is shaped by
both early adoption and latecomer effects.
\end{quote}

These two hypotheses are empirically tested in Section~\ref{sec-results}
employing the data and methods described in the next section.

\section{Data and Methods}\label{sec-datamethods}

To capture the diffusion of the commercial Web, a website density metric
is developed for two different geographical scales that are discussed
later in this section. The main data are counts of commercial websites
that are calculated using data from the Internet Archive\footnote{\url{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996--2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choice is
aligned with the topic of the paper -- the diffusion of the Web for
commercial functions -- and decreases the heterogeneity of the web data
as such commercial websites have specific aims: they are used to diffuse
information, support online transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as .com and these cases escape the data used
here, such omissions should not affect the validity of the results given
the popularity of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to
visit a .uk website when they are searching for products or services
\citep{hope}; and anecdotal evidence indicates that during the first
half of 2000, three .co.uk domains were registered every minute
\citep{oecd_coms}. Importantly, previous studies illustrated that .co.uk
is the most popular UK SLD \citep{tranosuk}. Henceforward any reference
to websites will refer to commercial -- .co.uk -- websites.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files \citep{ons}.

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This
website reconstruction process lend itself to the construction of
\emph{website} instead of \emph{webpage} density metrics. Websites tend
to represent specific organisations or entities, and, arguably, are more
meaningful observational units than webpages, which can ignore the
upstream dependency of the website they belong to. Based on the
following example, all three webpages are part of the same website
(\url{http://www.website.co.uk}). Only the overall website and not the
nested webpages are considered for the counts as otherwise the density
metrics would have been biased towards large, place specific, websites.
It is important to note here that websites commonly contain place
references, with previous research estimating this share to be around
70\% \citep{hill2009georeferencing}.

\begin{itemize}
\item
  \url{http://www.website.co.uk/webpage_a} B15 2TT
\item
  \url{http://www.website.co.uk/webpage_b} BS8 1TH
\item
  \url{http://www.website.co.uk/webpage_c} B15 2TT
\end{itemize}

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
\href{http://www.website.co.uk}{www.website.co.uk} includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of the number of
postcodes per website for 2000 is presented in Table \ref{f2000}, which
illustrates a wide range. At the left end of the distribution, there are
websites anchored to a unique location (72\% of all the reconstructed
websites in 2000), which may represent a small company with a single
trading location. At the right end, there are websites with thousands of
different postcodes. Considering the time period of the analysis, such
cases can represent directories which used to be popular in the
pre-search engines early times of the commercial Internet as well as
real estate websites \citep{tranosuk}. Previous research which used the
same methodology for a small subset of these data illustrates the
robustness of the above approach: its results matched in terms of
accuracy past rounds of research based on extensive qualitative work,
interviews and observational studies for the same study area
\citep{shoreditch}.

The analysis is firstly conducted on websites, which only contain one
unique postcode. As a robustness check, the analysis is replicated for
an extended subset of websites, which include up to 10 unique postcodes
to capture commercial websites which point to multiple locations. They
are geolocated by equally attaching them to these locations. This
extended sample includes 94 percent of all the archived websites in 2000
(see Table \ref{f2000}).

\begin{table}

\caption{Number of unique postcodes per .co.uk website, 2000.\label{f2000}}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Postcodes & F & F (\%) & cumulative F\\
\midrule
(0,1] & 41,596 & 0.718 & 0.718\\
(1,2] & 6,451 & 0.111 & 0.830\\
(2,10] & 6,163 & 0.106 & 0.936\\
(10,100] & 2,975 & 0.051 & 0.988\\
(100,1000] & 646 & 0.011 & 0.999\\
\addlinespace
(1000,10000] & 62 & 0.001 & 1.000\\
(10000,100000] & 4 & 0.000 & 1.000\\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textit{Source: } Tranos et al. 2021}\\
\end{tabular}
\end{table}

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms \citep{BL2013links}.
The website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1,000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As Section~\ref{sec-results}
illustrates, this led to a small increase of the model predictive power.

To empirically test the two hypotheses stated in
Section~\ref{sec-litreview}, the website density metric was calculated
for two different scales the Local Authority Districts (LAD) and the
Output Areas (OA). The former is the lowest administrative unit in the
UK and there are 374 LAD. The latter is the smallest census-based
geographical unit and there are c.~230,000 of them in the UK. To create
the website density metric, the yearly website counts at the LAD level
are standardised by the number of firms in LAD to avoid biases
associated with LAD hosting a large number of firms. Given that there is
no such readily available statistic for the OA, the actual OA level
counts are used. However, because of the the consistent spatial
definition of OA (they host 40-250 households),\footnote{\url{https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies}}
website counts in OA are interpreted as a density metric too.

These website density metrics are used for the different stages of the
analysis. Starting with the first hypothesis (H\textsubscript{1}), to
test whether the S-curve accurately describes the cumulative adoption of
the Web at a local scale, the following logistic function (Equation
\ref{eq:s}) is estimated for the whole of the UK and then for each LAD
separately:

\begin{align}
y = k /(1 + e^{-(t-t_{o})})\label{eq:s}
\end{align}

\noindent \(k\) is the asymptote or, in other words, the saturation
level, \(b\) the overall growth rate, and \(t_{0}\) the \emph{inflection
point} of maximum growth at \(k/2\), where the logistic function is
symmetrical \citep{wilson201281}. The \(t_0\) of each LAD is compared
against the \(t_0\) of the UK to delineate whether a LAD reached that
point faster or slower than the country average. This heterogeneity is
discussed further when testing the second hypothesis
(H\textsubscript{2}). Importantly, an accuracy criterion was imposed and
only LAD with \(R^2 > 0.9\) were included in this analysis. To estimate
Equation \ref{eq:s} a self-starting logistic growth model was employed
using the \texttt{nls} and \texttt{SSlogis} functions in \texttt{R}.
This logistic growth model depicts the initial period of slow diffusion
of a new technology, which is followed by the fast, exponential growth
stage to then eventually slow down and saturate
\citep{wilson201281, grubler1999dynamics}.

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=0.7\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-2-1.pdf}

}

\caption{\label{correct}Yearly website counts per postcode (top 1000)
before (top) and after (down) data imputation}

\end{figure}%

Next, exploratory spatial data analysis provides some first insights of
whether the other two drivers of spatial diffusion -- namely
neighbourhood and hierarchy -- underpin the diffusion of the Web in the
UK. To capture the former and following \citet{ding2010modeling} the
Moran's I and the Local Index of Spatial Association (LISA) are
estimated for the website density. To capture the hierarchy mechanism,
the Gini coefficient -- an established metric of inequality -- is
calculated. All the above are computed and plotted longitudinally both
for the LAD and the OA.

To formally test H\textsubscript{1} a modelling framework is developed
to assess the above diffusion mechanisms. The overarching aim is to
build a model that can test how well these three mechanisms can predict
website density:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S\text-curve_{t}\label{eq:rf.generic}
\end{align}

To capture the hierarchical mechanism the model includes as predictors a
one year lag of website density in London, the largest city in the UK,
and a one year lag of the website density in the nearest city. Both
variables clearly illustrate the urban hierarchy in the UK. Especially
for the OA models, a third predictor is added to capture hierarchy at a
much more granular scale: a one year lag of the website density in the
the nearest retail centre \footnote{The retail centres data have been
  obtained by the Consumer Data Research Centre, an ESRC data
  investment, under project ID CDRC 498-01, ES/L011840/1; ES/L011891/1:
  \url{https://data.cdrc.ac.uk/dataset/retail-centre-boundaries-and-open-indicators}.}.
As the literature illustrates, retail centres are key components of
urban systems \citep{dennis2002central, jones2021reframing}. Due to the
small sizes of the retail centres, the latter is only relevant for and
included in the OA models and not in the LAD models. By including this
predictor, the model is better tuned towards the spatial characteristic
of the OA scale. Hence, it increases the robustness of the model.

In addition, the model includes the distance to London, the nearest city
and the nearest retail centre. The underlying logic is that the level of
website adoption in a spatial unit depends on the level of the adoption
in places further up in the urban hierarchy the previous year. The
inclusion of the distance variables incorporates spatial structure into
the hierarchy argument. To depict the neighbourhood mechanism, the
website density of the neighbouring spatial units in the previous year
is employed. This is calculated using rook continuity. Again, the
underpinning rationale is that the level of web adoption within a
spatial unit depends on the level of web adoption in the neighbouring
spatial units the year before. This represents the `hitting nearby
locations first' argument. Therefore, the spatial and temporal lag of
the website density in LAD and OA is calculated. Lastly, the temporal
pattern of the cumulative adoption is captured by a time trend variable.
Hence, the model will follow the following generic form (Eq.
\ref{eq:model}):

\begin{align} \label{eq:model}
Website\,Density_{t} \sim Distance\,London +
Website\,density\,London_{t-1} +\notag\\
Distance\,Nearest\,City +
Website\,density\,Nearest\,City_{t-1} +\notag\\
Distance\,Nearest\,Retail_{i} +
Website\,density\,Nearest\,Retail_{t-1} +\notag\\
W*\, Website\,density_{t-1} +\notag\\ 
year_{t}
\end{align}

Specifically, four different models are estimated. Firstly, all the data
points for the OA and LAD are utilised in order to build two models and
assess their capacity to predict the adoption of the Web at the two
different scales. These two models will reveal the predictive capacity
of the diffusion mechanisms and also demonstrate how the importance of
such variables changes across different scales. The next two sets of
models are again trained at these two scales: OA and LAD. However,
instead of using all the data points, the OA and the LAD from one of the
twelve UK regions are held out for the model training. Then, the trained
model is used to predict website density for the OA or the LAD of the
held-out region. This process takes place recursively for all twelve UK
regions. The regional differences in the predictive capacity of the
different samples will reveal how dissimilar these spatial process are
across regions and, importantly, at different scales.

Equation \ref{eq:model} is estimated using Random Forest (RF). This is a
popular ML algorithm for both regression and classification problems
\citep{biau2012analysis}. It was introduced by \citet{breiman2001random}
and has become a go-to data science tool. RF can effectively handle
skewed distributions and outliers, model non-linear relationships,
require minimal hyperparameter tuning, exhibit low sensitivity to these
parameters, and have relatively short training times
\citep{Caruana2008, liaw2002classification, yan2020using}. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size (c.
230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors
\citep{mullainathan2017machine, athey2019machine, sulaiman2011intelligent, pourebrahim2019trip, biau2012analysis}.

RF is a tree-based ensemble learning algorithm
\citep{breiman2001random}. It begins by generating random samples of the
training data, which are then used to grow regression trees to predict
the dependent variable. Data points and predictors are randomly sampled
for the different trees. The trees are trained in parallel using their
own bootstrapped samples of the training data. A crucial feature of RF
is their ability not to overfit, meaning they can generalize well to
unseen test data. While each tree may overfit individually, the ensemble
of trees does not because the errors of individual trees are averaged,
reducing the overall variance and preventing overfitting
\citep{last2002improving}. For regression problems, RF predictions are
made by averaging the predictions of all decision trees.

RF have been widely employed to address regression research problems.
\citet{pourebrahim2019trip} combined a spatial interaction modelling
framework with ML algorithms including RF to predict commuting flows in
New York City. \citet{sinha2019assessing} advocated for adopting spatial
ensemble learning approaches, such as RF, to model spatial data with
high autocorrelation and heterogeneity. \citet{creditspatial} predicted
employment density in Los Angeles using spatially explicit RF.
\citet{guns2014recommending} used RF to build a recommendation system
for research collaborations. \citet{ren2019predicting} trained RF to
predict the socio-economic status of cities using various online and
mobility predictors. \citet{tranos2023using} utilised hyperlinks data
and RF to make out-of-sample predictions of interregional trade.
\citet{zhou2023geography} employed such a framework to assess whether
key predictors of obesity differ across English cities.

It needs to be highlighted here that the cross validation (CV) for all
models is spatially and temporally sensitive. Instead of using 10 random
and space- and time-agnostic folds, the \texttt{CAST} package is
employed which allows holding back data points from specific years and
spatial units (regions) and use them for testing in order to estimate
the model performance \citep{meyer2018improving}.

To assess the predictive power of the model, three broadly utilised
metrics are employed: the coefficient of determination (\(R^2\)), mean
absolute error (MAE) and root mean square error (RMSE):

\begin{align}
R^2 = 1 - \frac{\sum_{k} (y_{k} - \hat{y_{k}})^2} {\sum_{k} (y_{k} - \overline{y_{k}})^2} \label{eq:rsquared}
\end{align} \begin{align}
MAE = \frac{1}{N} \sum_{k = 1}^{N} |\hat{y_{k}} - y_{k}| \label{eq:mae}
\end{align} \begin{align}
RMSE =  \sqrt{\frac{\sum_{k = 1}^{N} (\hat{y_{k}} - y_{k})^2} {N}} \label{eq:rmse}
\end{align}

>>>>

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Economic Geography}

@@

\affiliation[1]{organization={University of Bristol, School of
Geographical Sciences},addressline={University Road, BS8
1SS},city={Bristol},country={UK},countrysep={,},postcodesep={}}

@@

\section{Acknowledgements}\label{acknowledgements}

Many thanks to Dr Levi Wolf and Professor Dani Arribas-Bel for their
constructive comments. Also many thanks to the participants of the 6th
Global Conference on Economic Geography and the Winter Seminar of GfR.
Then, many thanks to the organisers of the seminar series and the
members of the Quantitative Spatial Science (QuSS) Research Group at the
University of Bristol, the Centre for Business Prosperity Research Group
at Aston University (Dr Tasos Kitsos), and the Centre for Innovation
Research (Professor Tom Broekel) at the University of Stavanger.
Finally, many thanks to the editor and the three anonymous reviewers for
the very helpful comments and suggestions.

\section{Funding}\label{funding}

No external funding was received for this work.
